---
title: "Data Manipulation and Basic R"
author: "Jinwon Lee"
date: '2020 7 17 '
output:
  html_document:
    df_print: paged
---

###This material would be helpful to understand basic r: https://r4ds.had.co.nz 

#Packages 
Let's learn packages what usually used for data manipulation and wrangling. Here they are:
```{r,  warning= FALSE}
library(dplyr)
library(tidyverse)
library(haven)
library(readxl)
library(broom)

```

In turn:


*tidyverse*  includes:
1) ggplot2, which implements the Grammar of Graphics.

2) dplyr: One important contribution of the dplyr package is that it provides a “grammar” (in particular, verbs) for data manipulation and for operating on data frames. 

3) tidyr, which is a nice set of helper functions with an eye toward “tidy data:” a state of affairs where (1) every row in a data object is an observation; (2) every column in that data object is a variable; and (3) every cell in the data object is a single value.

These are major verbs of the tidyr : 

• select: return a subset of the columns of a data frame, using a flexible notation
• filter: extract a subset of rows from a data frame based on logical conditions
• arrange: reorder rows of a data frame
• rename: rename variables in a data frame
• mutate: add new variables/columns or transform existing variables
• summarise / summarize: generate summary statistics of different variables in the data frame,
possibly within strata
• %>%: the “pipe” operator is used to connect multiple verb actions together into a pipeline

4) readr is a good set of functions for reading in your data.

*haven* and *readxl* are both part of the tidyverse, too, but they are not automatically loaded. These help with reading in Stata, SAS, SPSS, and Excel data.

*broom* is a tidyverse-compliant way to handle the output of statistical models.

 

```{r}

print(table1)

data.frame(table1)

table1

# summarize(table1)

# learn functions : 1) filter

# see data collected in the year of 1999
filter(table1, year==1999) 

# see only 'China' data
filter(table1, country=='China')


#select? : see only country and year column 

select(table1, country, year)

# change the name of columm: rename

#e.g: rename(table1, numcases = cases)

#mutate

new_table1 <-mutate(table1, case_ratio = population/cases)

new_table1

```

### Download any dataset what you're interested in, and read that file by using different pakages (such as read.csv or readxl, etc).Then, try to manipulate data with what you've learned here. 




